Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-02 paper.tex
A few fixes following @najmahan's comments in https://github.com/glatard/FGCS-works-special-issue/pull/15/files

\title{Software Architectures for Integrating Workflow Engines in Science Gateways} \note{Naj}{'what' software architectures? what is their qualifier, adjective?}

@@

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. We describe and evaluate these software
  architectures with metrics that measure integration complexity,
  robustness, extensibility, scalability and other specific
  features\note{Naj}{Please be precise, what are they?}. Results illustrate the pros and cons of different
  architectures: Simple architectures with tight integration and task encapsulation are the
  the easiest to integrate and are most robust.Extensibility is equivalent in most architectures. The pool model is the most scalable one and meta-workflows are only available in nested workflows and workflow
  import. Our analysis provides insights for science gateway
  architects and workflow engine developers. \note{Naj}{I find this abstract to be vague, it can be made more concise and poignant}
\end{abstract}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools and
services offered by science gateways . Several
software architectures can be adopted to integrate workflow engines in
science gateways, however 
development effort required and the resulting system vary across different architectures. This paper and our analytical approach to comparison of software architectures is informed by our experince in the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years; as well as by lessons learned from several
science gateway and workflow projects, including
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
initiatives. This paper describes,
provides examples and compares such architectures, based on
system-independent representations of their main components and
interactions. In this section we provide background information and
definitions of different workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, different architectures are evaluated with
metrics measuring integration challenge, robustness, extensibility,
scalability and other specific features. We compare different architectures and highlight
recommendations for science gateway architects and workflow engine developers.

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
a high throughput. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM), \note{Naj}{should we not mention PMP which is used for CIVET?}
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

In this paper, the working definition of a
\emph{workflow engine} (also abbreviated \emph{engine}) is "a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values." Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution, and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications. \note{provide examples?}

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble other systems, for instance Apache
Airavata~\cite{marru2011apache}, the Catania Science Gateway
Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide},
and other ones, e.g. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13},
were built without any framework.

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}.  \closednote{Marc-e}{A
  few things bother me a bit with the notation fig., making it harder
  to assimilate 1- The interactions are padded in the middle of other
  stuff (Administrator, Data) 2- The Software component example says
  "Science gateway", should it not be empty?  3- The red arrow to
  explain the use of red; it looks like a very specific type of
  interactions instead of the desired "any interactions or components
  can be red meaning...", especially with 'a' and 'd' which the reader
  does not know the meaning yet.  Maybe a little red arrow and square
  with "Red used when specific to workflow execution"?  I don't have a
  great solution...  Red has special meaning, maybe remove it from the
  other pics (workflow eng, tasks) as it creates lots of visual
  interference when looking for red?  4- Order in text jumps all over
  the fig notations. Explanations are given for, in order: Components,
  workflow engine, interactions, abstract interactions, red color,
  tasks, data (no Administrator?)...  Could re-ordering the fig fix
  this? A figure figure legend could go a long way too (are there some
  in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}

Architectures are described from their main software \emph{components}
and \emph{interactions}. Software components include science gateway
and infrastructure in all architectures while workflow service,
workflow pool and agent are involved in some architectures only. The
workflow engine itself is represented by a specific symbol enclosed in a dotted rectangle.

Software interactions among these components are represented with
labeled, plain black arrows.  \emph{Abstract} interactions are
specific types of interactions that may be implemented using different
components and interactions. \note{Naj}{three 'interactions' in one sentence} They are represented with dotted arrows
and labeled with \texttt{*}. An architecture that has an abstract
interaction is an abstract architecture that can be instantiated in
various ways.

Red color (WHERE?) is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks executed on the
infrastructure. Data represents any type of file or database entry involved
in workflow execution and stored on the infrastructure. It does not
cover workflow parameters such as strings, numbers, etc. Administrator
refers to a science gateway user with extended privileges to install
applications. \note{Naj}{I propose a table of definitions.}

The remainder of this section describes the different types of
interactions involved in the architectures. Architectures are then
detailed and illustrated using examples of real
systems. Table~\ref{table:system-classification} summarizes the
classification of a few systems by architecture.

@@

\begin{figure}
\centering
\def\svgwidth{0.7\columnwidth}
\input{figures/notations.pdf_tex}
\caption{Graphical notations}
\label{fig:notations}
\end{figure}

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}].{ \def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent}
workflow executed by the \emph{parent} engine is itself a workflow,
called \emph{child} workflow, that is executed by a \emph{child}
engine. \note{Naj} {this reads funny} The parent and child engines might use different workflow
description languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through abstract
interaction \texttt{*$_1$}. The science gateway also communicates with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions \texttt{*$_4$}
and \texttt{*$_5$}. The parent engine communicates with the child
engine through abstract interaction \texttt{*$_6$}. Administrator
installs workflows through interactions \texttt{a$_1$} and
\texttt{a$_2$}.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \note{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

>>>>

\title{An analysis of existing software architectures\\ to integrate workflow engines in science gateways\closednote{Naj}{'what' software architectures? what is their qualifier, adjective?}{Tristan}{There is no qualifier except 'for integrating workflow engines in science gatewats'. They are just software architectures :) The point of the paper is to describe them properly. It's a general formulation to cover any kind of software architecture that is used to integrated workflow engines in science gateways. Does it make sense?}}

@@

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. We describe and evaluate these
  architectures with metrics that measure integration complexity,
  robustness, extensibility, scalability and support for
  meta-workflows and fine-grained debugging\closednote{Naj}{Please be
    precise, what are they?}{Tristan}{Done, thanks}. Results
  illustrate the pros and cons of different architectures: Tight
  integration and task encapsulation are the easiest to integrate and
  are most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. Our analysis provides insights for science gateway
  architects and workflow engine developers. \note{Naj}{I find this
    abstract to be vague, it can be made more concise and poignant}
\end{abstract}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, however development effort required and the resulting system
vary across different architectures.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of different
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration challenge, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and PMP
(Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical Parametric
Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), \closednote{Naj}{should we not mention PMP which is
  used for CIVET?}{Tristan}{Done, as discussed.}
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution, and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute
applications. \closednote{Naj}{provide examples?}{Tristan}{Examples
  are mentioned in the last paragraph of this section. Moved it next.}

Various science gateways have been developed, including complete
frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
as standalone systems~\cite{SHER-14,GLAT-13}. Most of these science
gateway frameworks and instances include one or several workflow
engines.

@@

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}. This section describes the different types of
interactions involved in the architectures. Architectures are then
detailed and illustrated using examples of real
systems. Table~\ref{table:system-classification} summarizes the
classification of a few systems by architecture.

\closednote{Marc-e}{A
  few things bother me a bit with the notation fig., making it harder
  to assimilate 1- The interactions are padded in the middle of other
  stuff (Administrator, Data) 2- The Software component example says
  "Science gateway", should it not be empty?  3- The red arrow to
  explain the use of red; it looks like a very specific type of
  interactions instead of the desired "any interactions or components
  can be red meaning...", especially with 'a' and 'd' which the reader
  does not know the meaning yet.  Maybe a little red arrow and square
  with "Red used when specific to workflow execution"?  I don't have a
  great solution...  Red has special meaning, maybe remove it from the
  other pics (workflow eng, tasks) as it creates lots of visual
  interference when looking for red?  4- Order in text jumps all over
  the fig notations. Explanations are given for, in order: Components,
  workflow engine, interactions, abstract interactions, red color,
  tasks, data (no Administrator?)...  Could re-ordering the fig fix
  this? A figure figure legend could go a long way too (are there some
  in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}
\closednote{Naj}{I propose a table of definitions.}{Tristan}{Added
  information in Figure 1 and removed the corresponding text.}

@@

\begin{figure}
\centering
\def\svgwidth{0.8\columnwidth}
\input{figures/notations.pdf_tex}
\caption{Graphical notations}
\label{fig:notations}
\end{figure}

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}.]{\def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex
}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), workflows include other
workflows that are executed by a \emph{child} engine. However, parent
and child engines might use different languages and might run on
different infrastructures. A parent workflow is
also called meta-workflow. The science gateway communicates with the
parent engine through abstract interaction \texttt{*$_1$}. The science
gateway also communicates with the infrastructure to transfer user
data through abstract interactions \texttt{*$_2$} and
\texttt{*$_3$}. Both workflow engines communicate with the
infrastructure through abstract interactions \texttt{*$_4$} and
\texttt{*$_5$}. The parent engine communicates with the child engine
through abstract interaction \texttt{*$_6$}. Administrator installs
workflows through interactions \texttt{a$_1$} and \texttt{a$_2$}.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}
